Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-03-11 Manuscript/manuscript.Rmd
manuscript edits

After suitably reducing the feature space, we performed individual random forest classification models specific to cell line, data resolution, resampling technique, and predictor type. We assessed each models' ability to correctly identify the genomic bins that overlap with flanked TAD boundaries using F1-Score, MCC, and balanced accuracy. For 10 kb data resoltion, it was found that models with distance-type predictors performed better according to all metrics, across each cell line, regardless of resampling technique used (Figure 4). This result was consistent across each of the other data resolutions as well (Supplementary Figure 4). Notably, we saw marked improvements in performance for models implemented on resampled datasets when compared to no resampling. Of the, resampling techniques considered, it was found that SMOTE resampled datasets, at the 10 kb resoltion, yielded the highest model performances across each metric. This result was, for the most part, consistent for the other data resolutions. Performances of randomly under-sampled datasets were found to be comparable to, if not better than (at some isolated resolutions; see supplementary), SMOTE. The steap drop off in performance for randomly over-sampled and no resampled datasets was most evident when assessing F1-scores and balanced accuracies. Measures of MCC were found to increase more steadily toward models implemented on SMOTE resampled datasets. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

<!--
## DISTANCE TYPE PREDICTORS OUTPERFORM OTHER PREDICTOR-TYPES

For each cell line and resolution, we balanced the data using each resampling technique and performed a series of random forest classification algorithms on the full features space (all combined predictor types), as well as on type-specific feature spaces (overlap counts, overlap predictors, and distances). PR-curves were plotted and compared for 10 kb resolutions (Figures 4-8). For all curves, the baseline threshold used to assign a performance to a random model is defined as the ratio of genomic bins that overlapped with a flanked TAD boundary divided by the total number of genomic bins for the test set, respective to cell line and resolution. It was found that for the 10 kb resolution data, models with distance type predictors had consistently higher AUPRC values compared to other predictor types, across both cell line and resampling technique. This result was observed for the other resolutions as well (Supplementary Figuers 9-23l; Supplementary Table 2).

Recursive feature elimination was used to establish the optimal number of distance-type features to include in downstream analyses. It was determined that, for 10 kb resolution data, near optimal performance was obtained for models with only 32 predictors, across each resampling technique and for each cell line (Figures 9-13). Results remained consistent across the other data resolutions (Supplementary Figures 24-38). Therefore, the feature space was reduced for subsequent models to the top 32 predictors, respective to cell line, resolution, and resampling technique. 
-->

>>>>

After suitably reducing the feature space, we performed individual random forest classification models specific to cell line, data resolution, resampling technique, and predictor type. We assessed each models' ability to correctly identify the genomic bins that overlap with flanked TAD boundaries using F1-Score, MCC, and balanced accuracy. For 10 kb data resoltion, it was found that models with distance-type predictors performed better according to all metrics, across each cell line, regardless of resampling technique used (Figure 4). This result was consistent across each of the other data resolutions as well (Supplementary Figure 4). Notably, we saw marked improvements in performance for models implemented on resampled datasets when compared to no resampling.

## MODELS IMPLEMENTED ON SMOTE RESAMPLED DATA PERFORMED BEST

Of the, resampling techniques considered, it was found that SMOTE resampled datasets, at the 10 kb resoltion, yielded the highest model performances across each metric. This result was, for the most part, consistent for the other data resolutions. Performances of randomly under-sampled datasets were found to be comparable to, if not better than (at some isolated resolutions; see supplementary), SMOTE. The steap drop off in performance for randomly over-sampled and no resampled datasets was most evident when assessing F1-scores and balanced accuracies. Measures of MCC were found to increase more steadily toward models implemented on SMOTE resampled datasets. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

We then directly compared models using SMOTE resampled datasets across each data resolution to see how performances changed as resolution increased. For each cell line, we found that model performances tended to decline as data resolution increased. Random forest models implemented on 10 kb, SMOTE resampled, data resolutions, with distance-type predictors were found to perform best according to MCC and balanced accuracy (Figure 5). There was a slight improvement in F1-Score for the 25 kb resolution on the GM12878 cell line. Performances from the K562 cell line degraded much faster as resolution increased, while the worst performing models for the GM12878 cell line were at the 50 kb resolution, followed by a slight improvement in performance at the 100 kb resolution. To be comprehensive we evaluated the performances across resolutions of models with distance-type predictors for the other resampling techniques and found the results to be largely consistent (Supplementary Figure 5).
